---
title: 'IBIO 831: Final Project'
output: html_document
date: "2023-04-20"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
rm(list=ls())
setwd("~/Documents/GitHub/IBIO_831_final/")

# Load Libraries


```

## How does microbial diversity influence ectoparasite load in painted turtles? 

Intro 

Painted turtles (Chrysemys picta) are a common freshwater turtle species 
in North America. As they inhabit rivers, lakes, and ponds they are often subject to aquatic 
ectoparasites, such as leeches, which serve as vectors for more harmful blood parasites. Recently there has been an uptick in research investigating the microbiome of painted turtles, focusing largely on quantifying bacterial phyla present in the gut. However, little is known about how the microbial diversity in these organisms may influence fitness, or how they interact with stressors in their environment, such as ectoparasites. In this study, we survey three populations of painted turtles, quantifying ectoparasite load, fecal microbiota diversity, and body size for three hundred individuals in order to determine if gut microbial diversity mediates susceptibility to ectoparasites. We predicted that turtles with higher gut microbial diversity would be less susceptible to ectoparasite predation, and consequently would have fewer attached leeches. 


```{r DATA SIMULATION, include=FALSE}




```

```{r DATA TRANSFORMATIONS}


```

```{r MODEL BUILD}
# NULL MODEL

# ecto ~ diversity

# ecto ~ body size 

# ecto ~ diversity + body size 

# ecto ~ diversity * body size

```

```{r MODEL COMPARISON}
# AIC MODEL COMPARISON



```

```{r MODEL EVALUATION, echo=FALSE}
# plot residuals of best-fit model here:


```
[Paragraph to discuss model testing and best-fit model]

```{r RESULTS - RAW DATA WITH MODEL PREDICTIONS, echo=FALSE}




```

```{r PREPARE EFFECT SIZES }



```
[ Placeholder to discuss results - using effect sizes]
